Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Good progress with Section 3, but all a bit in flux right now.

We describe the semantics of cooperative kernels
(Section~\ref{sec:semantics}), then summarise the important
differences between programming with regular vs.\ cooperative kernels
(Section~\ref{sec:programmingguidelines}) and the non-functional
properties that a developer of a cooperative kernel, and the
implementer of a scheduler for cooperative kernels, should strive for
(Section~\ref{sec:nonfunctional}).  The semantics of cooperative
kernels has been guided by the applications that we have studied
(described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
might have taken different, equally reasonable, semantic decisions (Section~\ref{sec:semanticalternatives}).
We conclude the section by discussing the backwards-compatible nature
of cooperative kernels with respect to blocking algorithms designed to
run on current GPU platforms
(Section~\ref{sec:backwardscompatibility}).

@@

We now explain the semantics of these language extensions, after which
we explain how the work stealing and graph traveral algorithms of
Section~\ref{sec:openclexamples} can be turned into cooperative kernels
that use the extensions.  We present the semantics intuitively, using English.
In Appendix~\ref{appendix:semantics} we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.

@@

\ADComment{Maybe refer back to motivating examples?}
Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, but a global barrier
is not provided as a primitive in OpenCL, CUDA or HSA.  This is
because, as discussed in Section~\ref{sec:intro}, an inter-workgroup barrier
requires fair scheduling of workgroups.

@@

\lstset{basicstyle=\tt}
\begin{lstlisting}
  $\globalbarrier()$;
  if($\getgroupid()$ == 0) $\offerfork$();
  $\globalbarrier()$;
  $\offerkill()$;
  $\globalbarrier()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

@@

\paragraph{Examples}

@@

\subsection{Programming Guidelines}\label{sec:programmingguidelines}

@@

\lstset{basicstyle=\tt}
\begin{lstlisting}
  if($\getgroupid{0}$ != 0) $\offerkill$();
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

>>>>

We describe the semantics of cooperative kernels
(Section~\ref{sec:semantics}) and summarise the important
differences between programming with regular vs.\ cooperative kernels
by showing how the motivating examples of \mysec\ref{sec:openclexamples} can be adapted to use cooperative kernels
(Section~\ref{sec:programmingguidelines}).
We then focus on the non-functional
properties that a developer of a cooperative kernel, and the
implementer of a scheduler for cooperative kernels, should strive for
(Section~\ref{sec:nonfunctional}).  The semantics of cooperative
kernels has been guided by the practical applications we have studied
(described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
might have taken different and also reasonable semantic decisions (Section~\ref{sec:semanticalternatives}).
We conclude the section by discussing the backwards-compatible nature
of cooperative kernels with respect to blocking algorithms designed to
run on current GPU platforms
(Section~\ref{sec:backwardscompatibility}).

@@

We now explain the semantics of these language extensions.  We present the semantics intuitively, using English.
In Appendix~\ref{appendix:semantics} (see supplementary material) we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.

@@

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, as illustrated in the graph traversal example of \mysec\ref{sec:openclexamples} and \myfig\ref{fig:graphsearch}.
Because a global barrier requires fair scheduling of workgroups it cannot be safely implemented in OpenCL, CUDA or HSA, and thus is not provided as a primitive in any of these programming models.

@@

\lstset{basicstyle=\tt,numbers=none}
\begin{lstlisting}
  $\globalbarrier()$;
  if($\getgroupid()$ == 0) $\offerfork$();
  $\globalbarrier()$;
  $\offerkill()$;
  $\globalbarrier()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt,numbers=left}

@@

\subsection{Programming With Coopeative Kernels}\label{sec:programmingguidelines}

\ADComment{Revise: we're folding this into the overview of the examples.}

\ADComment{Fold this in:}

Using coopeative kernels requires some programmer effort

\ADComment{Lifted:} , after which
we explain how the work stealing and graph traveral algorithms of
Section~\ref{sec:openclexamples} can be turned into cooperative kernels
that use the extensions.

@@

@@

\lstset{basicstyle=\tt,numbers=none}
\begin{lstlisting}
  if($\getgroupid{0}$ != 0) $\offerkill$();
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt,numbers=left}
